PAMELA: an annotation-based Java Modeling Framework

### 2021-02-11 ResponseLetter.tex

% Title:    A LaTeX Template For Responses To a Referees' Reports
% Author:   Petr Zemek <s3rvac@gmail.com>
% Homepage: https://blog.petrzemek.net/2016/07/17/latex-template-for-responses-to-referees-reports/
% License:  CC BY 4.0 (https://creativecommons.org/licenses/by/4.0/)
\documentclass[10pt]{article}

% Allow Unicode input (alternatively, you can use XeLaTeX or LuaLaTeX)
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{xcolor}

\usepackage{microtype,xparse,tcolorbox}
\newenvironment{reviewer-comment }{}{}
\tcbuselibrary{skins}
\tcolorboxenvironment{reviewer-comment }{empty,
  left = 1em, top = 1ex, bottom = 1ex,
  borderline west = {2pt} {0pt} {black!20},
}
\ExplSyntaxOn
\NewDocumentEnvironment {response} { +m O{black!20} } {
  \IfValueT {#1} {
    \begin{reviewer-comment~ }
      \setlength\parindent{2em}
      \noindent
      \ttfamily #1
    \end{reviewer-comment~ }
  }
  \par\noindent\ignorespaces\color{blue}
} { \bigskip\par }

@@

\bigskip
Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador and Antoine

\pagebreak

\Reviewer{\#1}
\begin{response}{They then briefly discuss implementation details, in a single page; and then discuss 'industrial use cases and experiments'.

Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
} Response to the first comment.
\end{response}
\begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}

We fully agree with the reviewer. We have expanded the related work section and compared our work with existing metamodeling and models@runtime approaches. Moreover, we have included a number of new references as suggested by Reviewer 2.

\end{response}

@@

\begin{response}{What would also be interesting is a general discussion on performance of your tool, gains by using your tool, ...
}
Response to n-ieme comment.
\end{response}

\begin{response}{Your reference list is incomplete and inconsistent. Formod or Formosa? When were URLs accessed (add date). DOIs are missing. Ref. 5 should have the ISoLA track name and editors added. Ref. 7 is missing publisher, journal, conference. Inconsistent use of first names, first names + initials, first initial.}
We have expanded our reference list and fixed all (detected) insconsistencies.
\end{response}

@@

\begin{response}{The paper highlights include "Contract programming with assertions [remove the plural] checking at runtime", yet that is not mentioned in the paper at all.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
PAMELA have been thoroughly tested. Indeed, the test suite (which is available in package XXX) reaches a coverage of XX\% of the code.
\end{response}

@@

\begin{response}{p10 l32+: very little detail or quantification on the tool's code base. Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
  Response to n-ieme comment.
\end{response}

\ToolComments{\#1}

\begin{response}{gives "git clone git@github.com:openflexo-team/pamela.git", which (even when using my own username) does not work. However, "git clone https://github.com/openflexo-team/pamela.git" immediately does. Why not state the latter?
}

We apologize for the inconvenience. We have replaced "git clone git@github.com:openflexo-team/pamela.git" for "git clone https://github.com/openflexo-team/pamela.git" as suggested.
\end{response}

### 2021-02-11 ResponseLetter.tex

\bigskip
Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador and Antoine

\pagebreak

\Reviewer{\#1}
\begin{response}{They then briefly discuss implementation details, in a single page; and then discuss 'industrial use cases and experiments'.

Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
} Response to the first comment.
\end{response}
\begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
We fully agree with the reviewer. We have expanded the related work section and compared our work with existing metamodeling and models@runtime approaches. Moreover, we have included a number of new references as suggested by Reviewer 2.
\end{response}

@@

\begin{response}{how to use the tool is not describe properly. It is not clear if the developers need to write all the annotations. If an application is built with PAMELA, what is the link between the interfaces and the concrete classes. Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
  Response to n-ieme comment.
\end{response}

\begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
  Response to n-ieme comment.
\end{response}

\begin{response}{section 2 is too abstract to be very useful}
  Response to n-ieme comment.
\end{response}

\begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
We have remove this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
\end{response}

\begin{response}{the partially implemented class concept is not clear.}
The partially implemented class concept is very similar to the default methods introduced in Java 8 (which were not available when PAMELA was first created) permitting a sort of multiple inheritance. We have clarified this in the paper.
\end{response}

@@

\begin{response}{the additional features seem relevant, but they are not described. Either the authors describe them a bit more or they remove the section.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 3.4 it is not clear who writes what and at which step of the development process.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 5, the use of PAMELA in the two projects is not clear.}
  Response to n-ieme comment.
\end{response}

\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec), work of Glenn Cavarle and others from the model @runtime community.}
We have extended the related work section in order to discuss the suggested related works (among others).
\end{response}

\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviours; and 2) maintaining a continuous cycle of metamodeling... These two points are not clear previously in the paper.
}
We have removed that part as it was indeed misleading and not discussed in that form previously in the paper. We have rephrased as follows:

PAMELA is not focused on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in [20]) but on:  1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel  may  be  built  incrementally  without  the  need  for  code  generation and thus,  avoiding round-tripping issues;  and 2) providing default implementations for frequently used abstractions in order to ease development.
\end{response}

@@

\ToolComments{\#2}

\begin{response}{The link in the paper is not clickable and copy-and-paste does not work (because of latex?) Link found: https://github.com/openflexo-team/pamela}
Thanks for pointing this out. We have modified the latex settings so that this link (and all the others throughout the paper) is now clickable.
\end{response}

### 2021-02-12 6-Experimentations.tex
Updates from Overleaf

@@

>>>>

\sg{
The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes. 159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}. 
}

@@

\sg{
An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.
}

\sg{
Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.
}
